---
title: "Project"
output:
  html_document: default
  pdf_document: default
date: "2022-12-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Does the water heating method depend on the number of household members?

\##<http://www.sthda.com/english/wiki/chi-square-test-of-independence-in-r#>:\~:text=The%20chi%2Dsquare%20test%20of,categories%20of%20the%20two%20variables. Creating frame from 2 data sets for analysis

```{r}
data <- hogar_epf_2019
heating_type <- data$FUENAGUA
members <- data$NMIEMB
frame_q1 <- data.frame(heating_type, members)
frame_q1
```

Clearing the frame_q1 from NA values

```{r}
frame_q1_clear <- na.omit(frame_q1)
frame_q1_clear <- frame_q1_clear[!grepl(9, frame_q1_clear$heating_type),]
frame_q1_clear <- frame_q1_clear[!grepl("b", frame_q1_clear$heating_type),]
frame_q1_clear <- frame_q1_clear[!grepl(0, frame_q1_clear$members),]
frame_q1_clear
```

Preparing matrix for plots

```{r}
matrix_q1 <- data.matrix(chisq$observed)
matrix_q1
```

Drawing plots (balloon and mosaic)

```{r}
library("graphics")
library("gplots")
q1_balloonplot <- balloonplot(t(matrix_q1), main="Xui", xlab="", ylab="",
                        label = FALSE, show.margins = FALSE)

q1_mosaicplot <- mosaicplot(matrix_q1, shade = TRUE, las = 100, main = "hui")
```

Chi-square test

```{r}
chisq <- chisq.test(frame_q1_clear$heating_type, frame_q1_clear$members)

```

Pearson residuals can be easily extracted from the output of the function chisq.test():

```{r}
Pearson_residuals <- round(chisq$residuals, 3)
```

Let's visualize Pearson residuals using the package corrplot:

```{r}
library(corrplot)
corrplot(chisq$residuals, is.cor = FALSE)
```

r is the residual of the cell

```{r}
contrib <- 100*chisq$residuals^2/chisq$statistic
round(contrib, 3)
```

Visualize the contribution

```{r}
q1_corrplot <- corrplot(contrib, is.cor = FALSE)
```

Access to the values returned by chisq.test() function

```{r}
#p-value
p-value <- chisq$p.value
#mean
q1_mean <- chisq$estimate
```
